Peaks Over Threshold for Bursty Time Series

### 2018-05-09 article_springer/article_springer.Rmd
\ref{ => \@ref(

In this article, we restrict ourselves to the uncoupled case, where
the two sequences $X(\ell, n)_{n \in \mathbb N}$
and $T(\ell, n)_{n \in \mathbb N}$ are independent.[^1]
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have similar
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory (POT approach); likewise, the CTRE model
interprets these observations as noise and filters them out.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis). The reciprocal slope is returned as the estimate of the tail 
exponent.
For instance, the top-left panel in Figure \ref{fig:QQ-plots} shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`. The dynamic QQ-estimator plot (top-center panel)
plots the tail exponent estimate for the largest values at cutoffs 
varying up to the 5th order statistic, and the region of stability at 
0.8 indicates a recommended estimate.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill or QQ-Plot estimator becomes. 
The bottom-left panel of Figure \ref{fig:QQ-plots}, for instance, shows the 
static QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). 
Even by looking at different cutoffs via the dynamic QQ-estimator plot 
(bottom-center panel) one hardly identifies $0.8$ as a clear candidate for 
a tail parameter estimate.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly, on a logarithmic scale 
(see Figure \ref{fig:QQ-plots}, right column).
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
(Figure \ref{fig:flare-diagnostics-2}, right panel)
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

@@

The Theorem in Section \ref{sec:scaling} implies that for a high 
threshold $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where $b(\,)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

@@

The above approach, though theoretically sound, benefits from the following
practical adjustments (compare with Figure \ref{fig:flares}):

@@

Figures \ref{fig:flare-diagnostics-1} and \ref{fig:flare-diagnostics-2} 
show the diagnostic plots for a 
minimum threshold chosen at the 200th order statistic. 
There is some residual autocorrelation for the sequence of threshold 
exceedance times that is not accounted for by the CTRE model. 
The fit with a Mittag-Leffler distribution ($\beta = 0.8$) is good, 
though there are signs that the power-law tail tapers off for very large 
inter-threshold crossing times. 
There is no apparent dependence between threshold exceedance times and 
event magnitudes seen in the copula plot.

@@

According to Figure \ref{fig:flares}, for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T_\ell \sim {\rm ML}(\beta, k^{1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not 
memoryless, and the probability density of the time $t$ until the next threshold 
crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{1/\beta} \sigma_0)$ is the probability density of 
${\rm ML}(\beta, k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability 
distribution shifts towards larger values for the next crossing
(see Figure \ref{fig:hazard}, left panel). 
The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a 
decreasing function for the Mittag-Leffler distribution. 
The closer $\beta$ is to $1$, the more the hazard rate mimics that of an 
exponential distribution (a constant function, see Figure \ref{fig:hazard}, 
right panel).

@@

To test our inference method via stability plots, we have simulated `r params$n` 
independent
waiting time and magnitude pairs $(W_k, J_k)$ (see upper panel in 
Figure \ref{fig:thresholdedBursty}). In order to have exact 
analytical values available for $\beta$ and $\sigma_0$, a distribution for 
$W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. 
If we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, 
then due to the stability property we have the *equality* of distribution
$W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, 
for $b(n) = n^{1/\beta}$. 
Using the parametrisation of @SamorodnitskyTaqqu, a few lines of 
calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) 
show that $D$ must have the stable distribution 
$S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is 
implemented in the R package `stabledist` by @stabledist.

@@

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$.
The distribution of $J_k$ is irrelevant for the inference on $\beta$ and 
$\sigma_0$ (we have chosen unit exponential random variables). 
Figure \ref{fig:sim} displays plots of $\hat \beta(\ell)$ and 
$\hat \sigma(\ell)$ vs. $k$; recall that $k$ is the index of the order
statistics of $J_k$ at which the threshold $\ell$ is placed. 
Dotted lines show 95% confidence intervals, which are derived from the 
asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR].
The dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$, 
showing that our inference method identifies the 
parameters correctly.

>>>>

In this article, we restrict ourselves to the uncoupled case, where
the two sequences $X(\ell, n)_{n \in \mathbb N}$
and $T(\ell, n)_{n \in \mathbb N}$ are independent.[^1]
Figure \@ref(fig:thresholdedBursty) shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have similar
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory (POT approach); likewise, the CTRE model
interprets these observations as noise and filters them out.

@@

Since the Mittag-Leffler distribution is heavy-tailed, many researchers 
would intuitively give the highest importance to the tail behaviour of the 
distribution, and 
estimate the exponent of the tail function with established methods such 
as the Hill estimator. The QQ-estimator [@Kratz96] is closely related to 
the Hill estimator and fits a 
least squares line through the logarithms of the ordered statistics 
(y-axis) and the corresponding quantiles of the exponential distribution 
(x-axis). The reciprocal slope is returned as the estimate of the tail 
exponent.
For instance, the top-left panel in Figure \@ref(fig:QQ-plots) shows the 
(static) QQ-estimator for the magnitudes of the solar flare data, returning 
a good fit to a Pareto distribution with tail parameter 
`r round(flares_exponent, 2)`. The dynamic QQ-estimator plot (top-center panel)
plots the tail exponent estimate for the largest values at cutoffs 
varying up to the 5th order statistic, and the region of stability at 
0.8 indicates a recommended estimate.

However, as described nicely by @Resnick97, the less 
similar a heavy-tailed distribution is to a Pareto distribution, the less
useful a Hill or QQ-Plot estimator becomes. 
The bottom-left panel of Figure \@ref(fig:QQ-plots), for instance, shows the 
static QQ-estimator plot for $2000$ draws from the Mittag-Leffler distribution with 
tail parameter $0.8$. The stretched exponential shape means 
that the Mittag-Leffler distribution has more probability near $0$ than 
the Pareto distribution, severely biasing the estimator downwards 
(`r round(ML_exponent, 2)`). 
Even by looking at different cutoffs via the dynamic QQ-estimator plot 
(bottom-center panel) one hardly identifies $0.8$ as a clear candidate for 
a tail parameter estimate.

@@

Hence if there is some prior expectation that the data are drawn from the 
Mittag-Leffler distribution
(as is the the case for threshold exceedance times),
then we recommend avoiding QQ-estimators and Hill plots altogether, 
and instead examining QQ-plots directly, on a logarithmic scale 
(see Figure \@ref(fig:QQ-plots), right column).
The scale parameter $\sigma$ is irrelevant for a QQ-Plot. 
The tail parameter $\beta$ can be estimated quickly via the log-moment 
method by @Cahoy2013, or via maximum likelihood. 
Both estimators are implemented in `MittagLeffleR` 
[@MittagLeffleR]. 
Since the exponential distribution is nested in the Mittag-Leffler 
family of distributions, a standard likelihood ratio test can be 
performed, with the exponential distribution as a null model against 
a Mittag-Leffler distribution as the alternative. 
As an example, the threshold crossing times for the solar flare dataset
(Figure \@ref(fig:flare-diagnostics-2), right panel)
yield a difference in deviance of 
$\approx$ `r round(2 * (mlmle_out$loglik - mlmle_out$null_loglik))`, 
which evaluates to a strongly significant $\chi^2_1$ test statistic 
(p-value = $10^{`r round(log10(mlmle_out$p_value))`}$) for the null
hypothesis $\beta = 1$.

@@

The Theorem in Section \@ref(sec:scaling) implies that for a high 
threshold $\ell$ we may 
approximate the distribution of $T(\ell)$ with an 
${\rm ML}(\beta, b(1/p))$ distribution, 
where $b(\,)$ varies regularly at $\infty$ with parameter $1/\beta$.
Building on the POT (Peaks Over Threshold) method, we propose the 
following estimation procedure for the distribution of $T(\ell)$:

@@

The above approach, though theoretically sound, benefits from the following
practical adjustments (compare with Figure \@ref(fig:flares)):

@@

Figures \@ref(fig:flare-diagnostics-1) and \@ref(fig:flare-diagnostics-2) 
show the diagnostic plots for a 
minimum threshold chosen at the 200th order statistic. 
There is some residual autocorrelation for the sequence of threshold 
exceedance times that is not accounted for by the CTRE model. 
The fit with a Mittag-Leffler distribution ($\beta = 0.8$) is good, 
though there are signs that the power-law tail tapers off for very large 
inter-threshold crossing times. 
There is no apparent dependence between threshold exceedance times and 
event magnitudes seen in the copula plot.

@@

According to Figure \@ref(fig:flares), for a threshold $\ell$ at the $k$-th order 
statistic, the fitted threshold exceedance time distribution is 
$$
T_\ell \sim {\rm ML}(\beta, k^{1/\beta} \sigma_0), 
$$
where $\beta = 0.85$ and $\sigma_0 = 3.0 \times 10^7 {\rm sec}$. 
Unlike the exponential distribution, the Mittag-Leffler distribution is not 
memoryless, and the probability density of the time $t$ until the next threshold 
crossing will depend on the time $t_0$ elapsed since the last threshold crossing.
This density equals 
$$
p(t|\beta, \sigma_0, \ell, t_0) = \frac{f(t + t_0 | \beta, k^{1/\beta} \sigma_0)}{\mathbf P[T_\ell > t_0]}
$$
where $f(\,\cdot\, | \beta, k^{1/\beta} \sigma_0)$ is the probability density of 
${\rm ML}(\beta, k^{1/\beta} \sigma_0)$. 
The more time has passed without a threshold crossing, the more the probability 
distribution shifts towards larger values for the next crossing
(see Figure \@ref(fig:hazard), left panel). 
The hazard rate 
$$
h(t) = \frac{f(t| \beta, k^{1/\beta} \sigma_0))}{\int_t^\infty f(\tau| \beta, k^{1/\beta} \sigma_0))\,d\tau}
$$
represents the risk of a threshold crossing per unit time, and is a 
decreasing function for the Mittag-Leffler distribution. 
The closer $\beta$ is to $1$, the more the hazard rate mimics that of an 
exponential distribution (a constant function, see Figure \@ref(fig:hazard), 
right panel).

@@

To test our inference method via stability plots, we have simulated `r params$n` 
independent
waiting time and magnitude pairs $(W_k, J_k)$ (see upper panel in 
Figure \@ref(fig:thresholdedBursty). In order to have exact 
analytical values available for $\beta$ and $\sigma_0$, a distribution for 
$W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. 
If we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, 
then due to the stability property we have the *equality* of distribution
$W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, 
for $b(n) = n^{1/\beta}$. 
Using the parametrisation of @SamorodnitskyTaqqu, a few lines of 
calculation (see e.g. the vignette on parametrisation in @MittagLeffleR) 
show that $D$ must have the stable distribution 
$S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is 
implemented in the R package `stabledist` by @stabledist.

@@

By the Theorem, the distribution 
of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$.
The distribution of $J_k$ is irrelevant for the inference on $\beta$ and 
$\sigma_0$ (we have chosen unit exponential random variables). 
Figure \@ref(fig:sim) displays plots of $\hat \beta(\ell)$ and 
$\hat \sigma(\ell)$ vs. $k$; recall that $k$ is the index of the order
statistics of $J_k$ at which the threshold $\ell$ is placed. 
Dotted lines show 95% confidence intervals, which are derived from the 
asymptotic normality of the log-moments estimators [@Cahoy2013] and the $\delta$-method [@MittagLeffleR].
The dashed lines show the actual values of $\beta$ resp.\ $\sigma_0$, 
showing that our inference method identifies the 
parameters correctly.
